---
title: "Reproducible Research: Peer Assessment 1"
author: "Sriram Pasupathi Raj"
date: "February 26, 2018"
output: 
  html_document: 
    keep_md: TRUE
---
```{r Global_options, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, highlight = TRUE, tidy = FALSE)
```

####LOAD AND PREPROCESS/TRANSFORM THE DATA INTO SUITABLE FORMAT FOR ANALYSIS

1. Load & Read the data using read.csv() setting the argument stringAsFactors = FALSE.
2. The date variable of the dataset is converted into 'Date' class (as a part of data transformation)
3. Output from  the below mentioned code chunk is a Tidy dataset because
    * Each measured variable is in one column
    * Each different observation of that variable are in a different row
    * There is only one table for the specified kind of variable
    
```{r Data_Preprocess, eval = TRUE, highlight = TRUE}
activity <- read.csv("Activity Monitoring Data/activity.csv", header = TRUE, stringsAsFactors = FALSE)
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")

#Dataset without Missing values (NAs)
activity_complete <- activity[complete.cases(activity), ]
head(activity_complete)
```

####MEAN AND MEDIAN OF THE TOTAL NUMBER OF STEPS TAKEN PER DAY

```{r Mean_&_Median}
Total_steps <- tapply(activity$steps, activity$date, sum, na.rm = TRUE)
print(mean(Total_steps))
print(median(Total_steps))
#Histogram of total number of steps taken each day
hist(Total_steps, col = "grey", xlab = "Total Number of steps per day", ylab = "Frequency", main = "Histogram of Total Number of steps taken each day")
dev.copy(png, file = "plot_1.png")
dev.off()
#Mean & Median of total number of steps taken each day without missing values
Mean <- aggregate(steps ~ date, activity_complete, mean)
Median <- aggregate(steps ~ date, activity_complete, median)
Mean_and_Median <- merge(Mean, Median, by.x = "date", by.y = "date", all = TRUE)
names(Mean_and_Median)[2:3] <- c("mean", "median")
print(Mean_and_Median)
```

####AVERAGE DAILY ACTIVITY PATTERN

```{r Daily_activity_pattern}
Total_interval <- aggregate(steps ~ interval, activity_complete, sum)
with(Total_interval, plot(interval, steps, type = "l", xlab = "5-minute interval", ylab = "Average number of steps taken, averaged across all days", main = "Time series plot of Average daily activity pattern", col = "dark red"))
dev.copy(png, file = "plot_2.png")
dev.off()
library(dplyr)
Maximum_steps <- filter(Total_interval, steps == max(Total_interval))
print(Maximum_steps)
```

####IMPUTING MISSING VALUES

```{r Imputing_missing_values}
Total_missing <- sum(is.na(activity))
print(Total_missing)

#My strategy for filling in all of the missing values in the dateset (Median of total number of steps taken, averaged across all days for each interval)
activity_NA <- activity[is.na(activity), ]
Median_interval <- tapply(activity_complete$steps, activity_complete$interval, median)
Activity_NA <- data.frame()
for(i in 1:length(Median_interval))
{
	Initial <- activity_NA[activity_NA$interval == names(Median_interval[i]), ]
	Initial$steps <- Median_interval[i]
	Activity_NA <- rbind(Activity_NA, Initial)
}
Activity <- rbind(Activity_NA, activity_complete)
Activity <- arrange(Activity, date)

#Histogram, Mean and Median of Total number of steps taken each day
New_Total_steps <- aggregate(steps ~ date, Activity, sum)
print(mean(New_Total_steps$steps))
print(median(New_Total_steps$steps))
hist(New_Total_steps$steps, col = "dark green", xlab = "Total number of steps taken per day", ylab = "Frequency", main = "Histogram of Total number of stpes taken each day")
dev.copy(png, file = "plot_3.png")
dev.off()

New_Mean_steps <- aggregate(steps ~ date, Activity, mean)
New_Median_steps <- aggregate(steps ~ date, Activity, median)
New_mean_and_median <- merge(New_Mean_steps, New_Median_steps, by.x = "date", by.y = "date", all = TRUE)
names(New_mean_and_median)[2:3] <- c("mean", "median")
print(New_mean_and_median)
```

####DIFFERENCES IN ACTIVITY PATTERNS BETWEEEN WEEKDAYS AND WEEKENDS

```{r Differences_between_weekdays_&_weekends}
library(lubridate)
Activity$days <- wday(Activity$date)
for(i in 1:nrow(Activity))
{
	if(Activity$days[i] == 1)
	{
		Activity$category[i] = "weekend"
	}
  else if(Activity$days[i] >= 2 && Activity$days[i] <= 6)
	{
		Activity$category[i] = "weekday"
	}
	else if(Activity$days[i] == 7)
	{
		Activity$category[i] = "weekend"
	}
}
Activity <- transform(Activity, category = factor(category))
Average_steps <- aggregate(steps ~ interval + category, Activity, mean)
library(lattice)
xyplot(steps ~ interval | category, data = Average_steps, type = "l", layout = c(1, 2))
dev.copy(png, file = "plot_4.png")
dev.off()
```